Supplementary Information part 3: Testing different forest cover relationships

### 2022-09-28 paper/Forest_and_Water/SupplementaryMaterialPart3.tex
revision of draft for internal review

In this document we tested the following, which is outlined in the methods of the paper.
The changes in forest cover contain both positive (forestation) and negative values (deforestation). In \citet{zhang2017}, these changes were jointly analysed, assuming the effect on the change in flow was linear and the effect of removing forest cover was the same as an equivalent addition of forest cover. Here we test two variation of this approach, leading to three approaches

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Joint analysis and assuming a linear relationship between the change in streamflow and the change in forest cover, following \citet{zhang2017}.
\item
  The impact of an increase in forest cover can be different from the same fractional decrease in forest cover. The question becomes how best to analyse this. One approach would be to allow a different slope and a different intercept for the decreases relative to the increases.
\item
  A second approach is to test the change in forest cover as a non-linear relationship in the Generalised additive model (GAM).
\end{enumerate}

@@

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.4231}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1410}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1667}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1282}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1410}}@{}}
\caption{(\#tab:m\_all-linear) Statistical summary for the linear terms the full model}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Std. Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
t value
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Pr(\textgreater\textbar t\textbar)
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Std. Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
t value
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Pr(\textgreater\textbar t\textbar)
\end{minipage} \\
\midrule()
\endhead
\textbf{(Intercept)} & -7.47 & 15.87 & -0.47 & 0.64 \\
\textbf{DeltaF\_perc} & -0.59 & 0.05 & -10.89 & 0 \\
\textbf{Precip\_data\_typeOB} & -17.87 & 12.91 & -1.38 & 0.17 \\
\textbf{Precip\_data\_typeSG} & 0.2 & 14.82 & 0.01 & 0.99 \\
\textbf{Assessment\_techniqueEA, HM} & 18.67 & 41.7 & 0.45 & 0.65 \\
\textbf{Assessment\_techniqueHM} & 26.61 & 11.43 & 2.33 & 0.02 \\
\textbf{Assessment\_techniquePWE} & 30.77 & 11.68 & 2.63 & 0.01 \\
\textbf{Assessment\_techniquePWE,
HM} & 15.8 & 42.22 & 0.37 & 0.71 \\
\textbf{Assessment\_techniqueQPW} & 41.35 & 19.66 & 2.1 & 0.04 \\
\textbf{Assessment\_techniqueQPW,
EA} & 26.05 & 23.84 & 1.09 & 0.28 \\
\textbf{Assessment\_techniqueSH} & 39.31 & 11.54 & 3.41 & 0 \\
\textbf{Forest\_typeCF} & -9.28 & 7.41 & -1.25 & 0.21 \\
\textbf{Forest\_typeMF} & -6.34 & 7.38 & -0.86 & 0.39 \\
\textbf{Hydrological\_regimeSD} & 0.13 & 8.94 & 0.01 & 0.99 \\
\bottomrule()
\end{longtable}

@@

For this model, the focus can be on the linear variables. The variable \texttt{DeltaF\_perc} clearly explains the variation in the change in flow with a p-value of 0. The overall model explains 50 \% of the variation in the data. For comparison across models with different structures and different variables, the Aikaike Information Criteria (AIC) can be used. The AIC of the model is 3166.6801428.

@@

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.4231}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1410}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1667}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1282}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1410}}@{}}
\caption{(\#tab:m\_all2-linear) Statistical summary for the linear terms the alternative model where the change is forestry is all positive}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Std. Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
t value
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Pr(\textgreater\textbar t\textbar)
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Std. Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
t value
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Pr(\textgreater\textbar t\textbar)
\end{minipage} \\
\midrule()
\endhead
\textbf{(Intercept)} & 10.91 & 19.47 & 0.56 & 0.58 \\
\textbf{DeltaF\_perc\_pos} & 0.29 & 0.1 & 2.97 & 0 \\
\textbf{Forest\_SignIncrease} & -52.26 & 7.62 & -6.86 & 0 \\
\textbf{Precip\_data\_typeOB} & -15.75 & 14.22 & -1.11 & 0.27 \\
\textbf{Precip\_data\_typeSG} & -5.2 & 16.81 & -0.31 & 0.76 \\
\textbf{Assessment\_techniqueEA, HM} & 14.21 & 45.49 & 0.31 & 0.76 \\
\textbf{Assessment\_techniqueHM} & 27 & 13.31 & 2.03 & 0.04 \\
\textbf{Assessment\_techniquePWE} & 28.19 & 13.57 & 2.08 & 0.04 \\
\textbf{Assessment\_techniquePWE,
HM} & 16.36 & 45.51 & 0.36 & 0.72 \\
\textbf{Assessment\_techniqueQPW} & 28.63 & 21.65 & 1.32 & 0.19 \\
\textbf{Assessment\_techniqueQPW,
EA} & 35.76 & 25.74 & 1.39 & 0.17 \\
\textbf{Assessment\_techniqueSH} & 33.52 & 12.94 & 2.59 & 0.01 \\
\textbf{Forest\_typeCF} & -12.47 & 8.06 & -1.55 & 0.12 \\
\textbf{Forest\_typeMF} & -5.1 & 8.18 & -0.62 & 0.53 \\
\textbf{Hydrological\_regimeSD} & 3.45 & 9.76 & 0.35 & 0.72 \\
\bottomrule()
\end{longtable}

Again the focus can be on the linear variables. In this case the variable \texttt{DeltaF\_perc\_pos} also is important to explain the variation in the change in flow with a p-value of 0. However, the results also show that in this case, the deforestation (\texttt{Forest\_Sign} \textless{} 0) has a different effect than reforestation (\texttt{Forest\_Sign} \textgreater{} 0). In fact, in absolute terms deforestration creates a -5226 \% greater effect on streamflow than reforestation (and of course deforestation creates and increase in flow and reforestation a decrease).

However, overall this model explains less of the overall variation in the data with a deviance explained of 0.46 and an AIC of 3212.700798.

@@

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3472}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0972}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1250}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1111}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}@{}}
\caption{(\#tab:m\_nonlin-smooth) Statistical summary for the smooth terms for the model that considers the change in forest cover to be a non-linear variable}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
edf
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Ref.df
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
F
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
p-value
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
edf
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Ref.df
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
F
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
p-value
\end{minipage} \\
\midrule()
\endhead
\textbf{s(DeltaF\_perc)} & 1.11 & 9 & 12.72 & 0 \\
\textbf{s(log10(Area\_km2))} & 0.76 & 4 & 0.8 & 0.03 \\
\textbf{s(Dryness)} & 4.7 & 9 & 2.2 & 0 \\
\textbf{s(length)} & 4.45 & 34 & 0.24 & 0.09 \\
\bottomrule()
\end{longtable}

In this case, the non-linear terms of the model need to be checked, which shows that once again, changes in forest cover impact changes in flow as indicated by the p-value of 0. The total deviance explained for the model is 0.5 and the model has an AIC of 3166.9317897. These performance statistics are similar to the first (baseline) model.

@@

\begin{verbatim}
## 
## Method: GCV   Optimizer: magic
## Smoothing parameter selection converged after 15 iterations.
## The RMS GCV score gradient at convergence was 0.004151441 .
## The Hessian was positive definite.
## Model rank =  69 / 69 
## 
## Basis dimension (k) checking results. Low p-value (k-index<1) may
## indicate that k is too low, especially if edf is close to k'.
## 
##                        k'    edf k-index p-value  
## s(DeltaF_perc)      9.000  1.113    0.89    0.03 *
## s(log10(Area_km2))  4.000  0.756    0.96    0.20  
## s(Dryness)          9.000  4.703    0.94    0.13  
## s(length)          34.000  4.448    0.92    0.10 .
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{verbatim}

>>>>

In this document we tested the following, which is outlined in the methods of the paper.
The changes in forest cover contain both positive (forestation) and negative values (deforestation). In \citet{zhang2017}, these changes were jointly analysed, assuming the effect on the change in flow was linear and the effect of removing forest cover was the same as an equivalent addition of forest cover. Here we test two variation of this approach, leading to three approaches:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Joint analysis and assuming a linear relationship between the change in streamflow and the change in forest cover, following \citet{zhang2017}.
\item
  The impact of an increase in forest cover can be different from the same fractional decrease in forest cover. The question becomes how best to analyse this. One approach would be to allow a different slope and a different intercept for the decreases relative to the increases.
\item
  A second approach is to test the change in forest cover as a non-linear relationship in the Generalised additive model (GAM).
\end{enumerate}

@@

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.4231}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1410}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1667}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1282}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1410}}@{}}
\caption{(\#tab:m\_all-linear) Statistical summary for the linear terms the full model}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Std. Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
t value
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Pr(\textgreater\textbar t\textbar)
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Std. Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
t value
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Pr(\textgreater\textbar t\textbar)
\end{minipage} \\
\midrule()
\endhead
\textbf{(Intercept)} & -4.81 & 16.31 & -0.29 & 0.77 \\
\textbf{DeltaF\_perc} & -0.6 & 0.06 & -10.71 & 0 \\
\textbf{Precip\_data\_typeOB} & -21.4 & 13.23 & -1.62 & 0.11 \\
\textbf{Precip\_data\_typeSG} & 9.36 & 15.17 & 0.62 & 0.54 \\
\textbf{Assessment\_techniqueEA, HM} & 20.64 & 42.73 & 0.48 & 0.63 \\
\textbf{Assessment\_techniqueHM} & 22.81 & 11.71 & 1.95 & 0.05 \\
\textbf{Assessment\_techniquePWE} & 30.63 & 11.94 & 2.57 & 0.01 \\
\textbf{Assessment\_techniquePWE,
HM} & 17.42 & 43.26 & 0.4 & 0.69 \\
\textbf{Assessment\_techniqueQPW} & 39.52 & 20.15 & 1.96 & 0.05 \\
\textbf{Assessment\_techniqueQPW,
EA} & 24.39 & 24.42 & 1 & 0.32 \\
\textbf{Assessment\_techniqueSH} & 45.3 & 11.83 & 3.83 & 0 \\
\textbf{Forest\_typeCF} & -9.45 & 7.6 & -1.24 & 0.21 \\
\textbf{Forest\_typeMF} & -8.05 & 7.56 & -1.06 & 0.29 \\
\textbf{Hydrological\_regimeSD} & 3.57 & 9.16 & 0.39 & 0.7 \\
\bottomrule()
\end{longtable}

@@

For this model, the focus can be on the linear variables. The variable \texttt{DeltaF\_perc} clearly explains the variation in the change in flow with a p-value of 0. The overall model explains 49 \% of the variation in the data. For comparison across models with different structures and different variables, the Aikaike Information Criteria (AIC) can be used. The AIC of the model is 3181.7600577.

@@

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.4231}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1410}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1667}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1282}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1410}}@{}}
\caption{(\#tab:m\_all2-linear) Statistical summary for the linear terms the alternative model where the change is forestry is all positive}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Std. Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
t value
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Pr(\textgreater\textbar t\textbar)
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Std. Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
t value
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Pr(\textgreater\textbar t\textbar)
\end{minipage} \\
\midrule()
\endhead
\textbf{(Intercept)} & 14.05 & 20 & 0.7 & 0.48 \\
\textbf{DeltaF\_perc\_pos} & 0.29 & 0.1 & 2.9 & 0 \\
\textbf{Forest\_SignIncrease} & -51.91 & 7.79 & -6.67 & 0 \\
\textbf{Precip\_data\_typeOB} & -19.43 & 14.56 & -1.33 & 0.18 \\
\textbf{Precip\_data\_typeSG} & 4.43 & 17.22 & 0.26 & 0.8 \\
\textbf{Assessment\_techniqueEA, HM} & 18.4 & 46.56 & 0.4 & 0.69 \\
\textbf{Assessment\_techniqueHM} & 21.99 & 13.63 & 1.61 & 0.11 \\
\textbf{Assessment\_techniquePWE} & 27.8 & 13.85 & 2.01 & 0.05 \\
\textbf{Assessment\_techniquePWE,
HM} & 17.58 & 46.6 & 0.38 & 0.71 \\
\textbf{Assessment\_techniqueQPW} & 25.71 & 22.18 & 1.16 & 0.25 \\
\textbf{Assessment\_techniqueQPW,
EA} & 33.3 & 26.36 & 1.26 & 0.21 \\
\textbf{Assessment\_techniqueSH} & 38.7 & 13.25 & 2.92 & 0 \\
\textbf{Forest\_typeCF} & -12.71 & 8.25 & -1.54 & 0.12 \\
\textbf{Forest\_typeMF} & -7.05 & 8.36 & -0.84 & 0.4 \\
\textbf{Hydrological\_regimeSD} & 7.27 & 10 & 0.73 & 0.47 \\
\bottomrule()
\end{longtable}

Again the focus can be on the linear variables. In this case the variable \texttt{DeltaF\_perc\_pos} also is important to explain the variation in the change in flow with a p-value of 0. However, the results also show that in this case, the deforestation (\texttt{Forest\_Sign} \textless{} 0) has a different effect than reforestation (\texttt{Forest\_Sign} \textgreater{} 0). In fact, in absolute terms deforestration creates a -5191 \% greater effect on streamflow than reforestation (and of course deforestation creates and increase in flow and reforestation a decrease).

However, overall this model explains less of the overall variation in the data with a deviance explained of 0.45 and an AIC of 3227.3976077.

@@

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3472}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0972}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1250}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0972}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}@{}}
\caption{(\#tab:m\_nonlin-smooth) Statistical summary for the smooth terms for the model that considers the change in forest cover to be a non-linear variable}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
edf
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Ref.df
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
F
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
p-value
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
edf
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Ref.df
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
F
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
p-value
\end{minipage} \\
\midrule()
\endhead
\textbf{s(DeltaF\_perc)} & 4.14 & 9 & 13.2 & 0 \\
\textbf{s(log10(Area\_km2))} & 0.74 & 4 & 0.74 & 0.04 \\
\textbf{s(Dryness)} & 4.73 & 9 & 2.4 & 0 \\
\textbf{s(length)} & 4.68 & 34 & 0.26 & 0.08 \\
\bottomrule()
\end{longtable}

In this case, the non-linear terms of the model need to be checked, which shows that once again, changes in forest cover impact changes in flow as indicated by the p-value of 0. The total deviance explained for the model is 0.5 and the model has an AIC of 3182.1088731. These performance statistics are similar to the first (baseline) model.

@@

\begin{verbatim}
## 
## Method: GCV   Optimizer: magic
## Smoothing parameter selection converged after 11 iterations.
## The RMS GCV score gradient at convergence was 0.0003791486 .
## The Hessian was positive definite.
## Model rank =  69 / 69 
## 
## Basis dimension (k) checking results. Low p-value (k-index<1) may
## indicate that k is too low, especially if edf is close to k'.
## 
##                       k'   edf k-index p-value  
## s(DeltaF_perc)      9.00  4.14    0.91   0.075 .
## s(log10(Area_km2))  4.00  0.74    0.95   0.140  
## s(Dryness)          9.00  4.73    0.93   0.135  
## s(length)          34.00  4.68    0.93   0.110  
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{verbatim}
